---
title: 'Historical Data Analysis of Storm Effects on Human Health and Economy in United States'
subtitle: 'Coursera Reproducible Research Assignment #2'
author: "Peter EerNisse - https://github.com/peernisse/5_Week_4_RItems"
date: "September 14, 2018"
output: 

        html_document:
                toc: TRUE
                toc_float: TRUE
                code_folding: hide
---

```{r setup, include=TRUE, echo=TRUE, warning= FALSE, results='hide'}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE,include=TRUE,error = FALSE)

```

#Synopsis

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project explored the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm databaseto estimate likelihood of fatalities, injuries, and property damage given a certain storm event type.

This analysis addresses the following objectives/questions:

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

In summary, human health, as assessed by injury and fatality data, were most likely in cases of tsunami, and extreme weather events such as fog and stormy/icy weather, as well as extreme heat or extreme cold. Property damage was highest in events of hurricane, severe storms, drought and floods.

The analysis process and results are presented in the following sections.

#Data Processing

Data and supporting information for the analysis are sourced from the National Oceanic and Atmoshperic Administration, provided as course materials by Coursera, and were downloaded from the following links: 

* [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)
* [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
* [National Climatic Data Center Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

##Data Import
The raw data were imported in R using the RStudio text editor. Review the code chunk below to see the download code. 

<!--Data Import-->
```{r data import, echo=TRUE, results='hide', cache=TRUE}

#Libraries
library(tidyverse)
library(knitr)

#Connect and download data file
#One time only to get csv file uncomment to run
download.file('https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2',destfile = './data/rawData.csv')

#One time only uncomment to run
rawData<-read.csv('./data/rawData.csv',stringsAsFactors = FALSE)

```

##Data Cleanup

Data were processed as follows:

1. Separate injury and fatality data from property damage data.
2. Remove any event types (EVTYPE) that did not result in at least one injury, fatality, or property damage.
3. Summarize the totals of results by event type, and transform property damage numbers into dollars based on the multiplier column PROPDMGEXP.
4. Remove any records with only a single event for that event type.
5. Manually re-group similar event types to address differences in spelleg and context (i.e., consolidate duplicate information into common category names)
6. Calculate the relative likelihood of injury or fatality as the number of recorded incidents divided by the number of weather events in that event type.
7. Calculate average property damage in dollars as the number of dollars divided by the number of weather events in that event type.

###Data reduction, calculations, and category consolidation. 


```{r data cleanup, cache=TRUE}
#Reduce data source for analysis

#Fix dates
newdates<-strptime(rawData$BGN_DATE, format = '%m/%d/%Y')
rawData$DATE<-as.POSIXct(newdates)

analData<-rawData %>% 
        select(STATE,EVTYPE,DATE,FATALITIES,INJURIES,PROPDMG,PROPDMGEXP) %>% 
        filter(FATALITIES > 0 | INJURIES > 0 | PROPDMG > 0, PROPDMGEXP %in% c('K','B','M','m')) %>% 
        mutate(DamageDollars = case_when(
             PROPDMGEXP == 'K' ~    PROPDMG*1000,
             PROPDMGEXP == 'M' ~    PROPDMG*1000000,
             PROPDMGEXP == 'm' ~    PROPDMG*1000000,
             PROPDMGEXP == 'B' ~    PROPDMG*1000000000
             
        )) %>% 
        
        group_by(EVTYPE) %>% 
        summarise(MinDate = min(DATE), MaxDate = max(DATE),NumberofEvents = length(DATE),SumFatality = sum(FATALITIES), SumInjury = sum(INJURIES), SumPropertyDamage = sum(DamageDollars))

#Separate human health and property damage data and remove if no events, or if number of events is 1
hhData<-analData %>% 
        select(EVTYPE,MinDate,MaxDate,NumberofEvents,SumFatality,SumInjury) %>% 
        filter(SumFatality > 0 | SumInjury > 0, NumberofEvents > 1)

pdData<-analData %>% 
        select(EVTYPE,MinDate,MaxDate,NumberofEvents,SumPropertyDamage) %>% 
        filter(SumPropertyDamage > 0 , NumberofEvents > 1)

#Create unique list of event types to build lookup
lookup<-rbind(hhData[,c(1,4)], pdData[,c(1,4)]) 
lookup<-unique(arrange(lookup,EVTYPE))

#Create lookup merge new event tyoe column to merge back into data frames
lookup<-lookup %>% 
        mutate(newEventType = case_when(
                EVTYPE == ' TSTM WIND' ~ 'WIND',
                EVTYPE == 'BRUSH FIRE' ~ 'WILDFIRE',
                EVTYPE == 'Coastal Flood' ~ 'COASTAL FLOOD',
                EVTYPE == 'Coastal Flooding' ~ 'COASTAL FLOOD',
                EVTYPE == 'COASTAL FLOODING' ~ 'COASTAL FLOOD',
                EVTYPE == 'COASTAL FLOODING/EROSION' ~ 'COASTAL FLOOD',
                EVTYPE == 'Cold' ~ 'EXTREME COLD',
                EVTYPE == 'COLD/WIND CHILL' ~ 'EXTREME COLD',
                EVTYPE == 'DRY MICROBURST' ~ 'DUST STORM',
                EVTYPE == 'Dust Devil' ~ 'DUST STORM',
                EVTYPE == 'DUST DEVIL' ~ 'DUST STORM',
                EVTYPE == 'Erosion/Cstl Flood' ~ 'COASTAL FLOOD',
                EVTYPE == 'EXCESSIVE HEAT' ~ 'EXTREME HEAT',
                EVTYPE == 'EXTREME COLD/WIND CHILL' ~ 'EXTREME COLD',
                EVTYPE == 'EXTREME WINDCHILL' ~ 'EXTREME COLD',
                EVTYPE == 'EXCESSIVE SNOW' ~ 'BLIZZARD',
                EVTYPE == 'FLASH FLOOD - HEAVY RAIN' ~ 'FLASH FLOOD',
                EVTYPE == 'FLASH FLOOD FROM ICE JAMS' ~ 'FLASH FLOOD',
                EVTYPE == 'FLASH FLOOD/FLOOD' ~ 'FLASH FLOOD',
                EVTYPE == 'FLASH FLOODING' ~ 'FLASH FLOOD',
                EVTYPE == 'FLASH FLOODING/FLOOD' ~ 'FLASH FLOOD',
                EVTYPE == 'FLASH FLOODS' ~ 'FLASH FLOOD',
                EVTYPE == 'FLOOD/FLASH FLOOD' ~ 'FLASH FLOOD',
                EVTYPE == 'FLOODING' ~ 'FLOOD',
                EVTYPE == 'FLOODS' ~ 'FLOOD',
                EVTYPE == 'DENSE FOG' ~ 'FOG',
                EVTYPE == 'FREEZING FOG' ~ 'FOG',
                EVTYPE == 'FREEZE' ~ 'ICE STORM',
                EVTYPE == 'Freezing Drizzle' ~ 'ICE STORM',
                EVTYPE == 'FREEZING DRIZZLE' ~ 'ICE STORM',
                EVTYPE == 'FREEZING RAIN' ~ 'ICE STORM',
                EVTYPE == 'FREEZING RAIN/SNOW' ~ 'ICE STORM',
                EVTYPE == 'FROST/FREEZE' ~ 'ICE STORM',
                EVTYPE == 'FUNNEL CLOUD' ~ 'TORNADO',
                EVTYPE == 'Glaze' ~ 'ICE STORM',
                EVTYPE == 'GLAZE' ~ 'ICE STORM',
                EVTYPE == 'GLAZE ICE' ~ 'ICE STORM',
                EVTYPE == 'gradient wind' ~ 'WIND',
                EVTYPE == 'Gradient wind' ~ 'WIND',
                EVTYPE == 'GROUND BLIZZARD' ~ 'BLIZZARD',
                EVTYPE == 'GUSTNADO' ~ 'TORNADO',
                EVTYPE == 'GUSTY WIND' ~ 'WIND',
                EVTYPE == 'Gusty Winds' ~ 'WIND',
                EVTYPE == 'GUSTY WINDS' ~ 'WIND',
                EVTYPE == 'HAIL 175' ~ 'HAIL',
                EVTYPE == 'HAIL 275' ~ 'HAIL',
                EVTYPE == 'HAILSTORM' ~ 'HAIL',
                EVTYPE == 'HEAT' ~ 'EXTREME HEAT',
                EVTYPE == 'HEAT WAVE' ~ 'EXTREME HEAT',
                EVTYPE == 'HEAVY MIX' ~ 'RAIN',
                EVTYPE == 'HEAVY RAINS/FLOODING' ~ 'RAIN',
                EVTYPE == 'HEAVY SNOW' ~ 'HEAVY SNOW',
                EVTYPE == 'HEAVY SNOW-SQUALLS' ~ 'BLIZZARD',
                EVTYPE == 'HEAVY SNOW SQUALLS' ~ 'BLIZZARD',
                EVTYPE == 'HEAVY SNOW/ICE' ~ 'BLIZZARD',
                EVTYPE == 'HEAVY SNOW/SQUALLS' ~ 'BLIZZARD',
                EVTYPE == 'HEAVY RAIN' ~ 'RAIN',
                EVTYPE == 'HEAVY RAINS' ~ 'RAIN',
                EVTYPE == 'HEAVY SURF/HIGH SURF' ~ 'HEAVY SURF',
                EVTYPE == 'HIGH SEAS' ~ 'HEAVY SURF',
                EVTYPE == 'High Surf' ~ 'HEAVY SURF',
                EVTYPE == 'HIGH SURF' ~ 'HEAVY SURF',
                EVTYPE == 'HIGH WATER' ~ 'FLOOD',
                EVTYPE == 'HIGH WIND' ~ 'WIND',
                EVTYPE == 'HIGH WIND (G40)' ~ 'WIND',
                EVTYPE == 'HIGH WIND DAMAGE' ~ 'WIND',
                EVTYPE == 'HIGH WINDS' ~ 'WIND',
                EVTYPE == 'HIGH WINDS/COLD' ~ 'BLIZZARD',
                EVTYPE == 'HIGH WINDS/SNOW' ~ 'BLIZZARD',
                EVTYPE == 'HURRICANE-GENERATED SWELLS' ~ 'HURRICANE',
                EVTYPE == 'HURRICANE ERIN' ~ 'HURRICANE',
                EVTYPE == 'HURRICANE OPAL' ~ 'HURRICANE',
                EVTYPE == 'HURRICANE/TYPHOON' ~ 'HURRICANE',
                EVTYPE == 'ICE' ~ 'ICE STORM',
                EVTYPE == 'ICE FLOES' ~ 'FLOOD',
                EVTYPE == 'ICE JAM FLOODING' ~ 'FLOOD',
                EVTYPE == 'ICE STORM' ~ 'ICE STORM',
                EVTYPE == 'ICY ROADS' ~ 'ICE STORM',
                EVTYPE == 'LAKE-EFFECT SNOW' ~ 'HEAVY SNOW',
                EVTYPE == 'LAKE EFFECT SNOW' ~ 'HEAVY SNOW',
                EVTYPE == 'LAKESHORE FLOOD' ~ 'FLOOD',
                EVTYPE == 'LANDSLIDE' ~ 'LANDSLIDE',
                EVTYPE == 'LANDSLIDES' ~ 'LANDSLIDE',
                EVTYPE == 'LANDSPOUT' ~ 'TORNADO',
                EVTYPE == 'LIGHT SNOW' ~ 'HEAVY SNOW',
                EVTYPE == 'LIGHTNING' ~ 'LIGHTNING',
                EVTYPE == 'MAJOR FLOOD' ~ 'FLOOD',
                EVTYPE == 'MARINE HAIL' ~ 'HAIL',
                EVTYPE == 'LIGHT FREEZING RAIN' ~ 'ICE STORM',
                EVTYPE == 'Light Snow' ~ 'HEAVY SNOW',
                EVTYPE == 'MARINE HIGH WIND' ~ 'WIND',
                EVTYPE == 'MARINE STRONG WIND' ~ 'WIND',
                EVTYPE == 'MARINE THUNDERSTORM WIND' ~ 'WIND',
                EVTYPE == 'MARINE TSTM WIND' ~ 'WIND',
                EVTYPE == 'MICROBURST' ~ 'WIND',
                EVTYPE == 'Mixed Precipitation' ~ 'RAIN',
                EVTYPE == 'MIXED PRECIPITATION' ~ 'RAIN',
                EVTYPE == 'MUD SLIDE' ~ 'LANDSLIDE',
                EVTYPE == 'MUDSLIDE' ~ 'LANDSLIDE',
                EVTYPE == 'OTHER' ~ 'OTHER',
                EVTYPE == 'RAIN' ~ 'RAIN',
                EVTYPE == 'RECORD COLD' ~ 'EXTREME COLD',
                EVTYPE == 'RECORD SNOW' ~ 'HEAVY SNOW',
                EVTYPE == 'RIP CURRENT' ~ 'HEAVY SURF',
                EVTYPE == 'RIP CURRENTS' ~ 'HEAVY SURF',
                EVTYPE == 'RIVER AND STREAM FLOOD' ~ 'FLOOD',
                EVTYPE == 'RIVER FLOOD' ~ 'FLOOD',
                EVTYPE == 'River Flooding' ~ 'FLOOD',
                EVTYPE == 'RIVER FLOODING' ~ 'FLOOD',
                EVTYPE == 'SEICHE' ~ 'FLOOD',
                EVTYPE == 'SEVERE THUNDERSTORM' ~ 'THUNDERSTORM',
                EVTYPE == 'SEVERE THUNDERSTORM WINDS' ~ 'WIND',
                EVTYPE == 'SEVERE THUNDERSTORMS' ~ 'THUNDERSTORM',
                EVTYPE == 'SMALL HAIL' ~ 'HAIL',
                EVTYPE == 'Snow' ~ 'HEAVY SNOW',
                EVTYPE == 'SNOW' ~ 'HEAVY SNOW',
                EVTYPE == 'SNOW AND HEAVY SNOW' ~ 'HEAVY SNOW',
                EVTYPE == 'SNOW FREEZING RAIN' ~ 'BLIZZARD',
                EVTYPE == 'SNOW SQUALL' ~ 'BLIZZARD',
                EVTYPE == 'Snow Squalls' ~ 'BLIZZARD',
                EVTYPE == 'SNOW SQUALLS' ~ 'BLIZZARD',
                EVTYPE == 'SNOW/COLD' ~ 'BLIZZARD',
                EVTYPE == 'SNOW/FREEZING RAIN' ~ 'BLIZZARD',
                EVTYPE == 'SNOW/HIGH WINDS' ~ 'BLIZZARD',
                EVTYPE == 'SNOW/ICE STORM' ~ 'ICE STORM',
                EVTYPE == 'SNOW/SLEET' ~ 'BLIZZARD',
                EVTYPE == 'SNOW/SLEET/FREEZING RAIN' ~ 'ICE STORM',
                EVTYPE == 'SNOWMELT FLOODING' ~ 'FLOOD',
                EVTYPE == 'STORM SURGE' ~ 'FLOOD',
                EVTYPE == 'STORM SURGE/TIDE' ~ 'FLOOD',
                EVTYPE == 'Strong Wind' ~ 'WIND',
                EVTYPE == 'STRONG WIND' ~ 'WIND',
                EVTYPE == 'STRONG WINDS' ~ 'WIND',
                EVTYPE == 'THUNDERSTORM' ~ 'THUNDERSTORM',
                EVTYPE == 'THUNDERSTORM  WINDS' ~ 'WIND',
                EVTYPE == 'THUNDERSTORM WIND' ~ 'WIND',
                EVTYPE == 'THUNDERSTORM WIND 60 MPH' ~ 'WIND',
                EVTYPE == 'THUNDERSTORM WIND/ TREES' ~ 'WIND',
                EVTYPE == 'THUNDERSTORM WINDS' ~ 'WIND',
                EVTYPE == 'THUNDERSTORM WINDS HAIL' ~ 'HAIL',
                EVTYPE == 'THUNDERSTORM WINDS LIGHTNING' ~ 'LIGHTNING',
                EVTYPE == 'THUNDERSTORM WINDS/ FLOOD' ~ 'FLOOD',
                EVTYPE == 'THUNDERSTORM WINDS/HAIL' ~ 'HAIL',
                EVTYPE == 'THUNDERSTORM WINDSS' ~ 'WIND',
                EVTYPE == 'THUNDERSTORMS WIND' ~ 'WIND',
                EVTYPE == 'THUNDERSTORMS WINDS' ~ 'WIND',
                EVTYPE == 'THUNDERTORM WINDS' ~ 'WIND',
                EVTYPE == 'TIDAL FLOODING' ~ 'COASTAL FLOOD',
                EVTYPE == 'TORNADO' ~ 'TORNADO',
                EVTYPE == 'TORNADO F0' ~ 'TORNADO',
                EVTYPE == 'TORNADO F1' ~ 'TORNADO',
                EVTYPE == 'TORNADO F2' ~ 'TORNADO',
                EVTYPE == 'TORNADO F3' ~ 'TORNADO',
                EVTYPE == 'TROPICAL DEPRESSION' ~ 'HURRICANE',
                EVTYPE == 'TROPICAL STORM' ~ 'HURRICANE',
                EVTYPE == 'TROPICAL STORM JERRY' ~ 'HURRICANE',
                EVTYPE == 'TSTM WIND' ~ 'WIND',
                EVTYPE == 'TSTM WIND (G40)' ~ 'WIND',
                EVTYPE == 'TSTM WIND (G45)' ~ 'WIND',
                EVTYPE == 'TSTM WIND 55' ~ 'WIND',
                EVTYPE == 'TSTM WIND/HAIL' ~ 'HAIL',
                EVTYPE == 'TSTM WINDS' ~ 'WIND',
                EVTYPE == 'TSUNAMI' ~ 'TSUNAMI',
                EVTYPE == 'TYPHOON' ~ 'HURRICANE',
                EVTYPE == 'URBAN FLOOD' ~ 'FLOOD',
                EVTYPE == 'URBAN FLOODING' ~ 'FLOOD',
                EVTYPE == 'URBAN FLOODS' ~ 'FLOOD',
                EVTYPE == 'URBAN/SMALL STREAM FLOOD' ~ 'FLOOD',
                EVTYPE == 'URBAN/SML STREAM FLD' ~ 'FLOOD',
                EVTYPE == 'VOLCANIC ASH' ~ 'VOLCANIC ASH',
                EVTYPE == 'WATERSPOUT' ~ 'WATERSPOUT',
                EVTYPE == 'WATERSPOUT-' ~ 'WATERSPOUT',
                EVTYPE == 'WATERSPOUT/TORNADO' ~ 'WATERSPOUT',
                EVTYPE == 'WET MICROBURST' ~ 'WET MICROBURST',
                EVTYPE == 'WILD FIRES' ~ 'WILDFIRE',
                EVTYPE == 'WILD/FOREST FIRE' ~ 'WILDFIRE',
                EVTYPE == 'WILDFIRE' ~ 'WILDFIRE',
                EVTYPE == 'WILDFIRES' ~ 'WILDFIRE',
                EVTYPE == 'WIND' ~ 'WIND',
                EVTYPE == 'WIND DAMAGE' ~ 'WIND',
                EVTYPE == 'WINDS' ~ 'WIND',
                EVTYPE == 'WINTER STORM' ~ 'BLIZZARD',
                EVTYPE == 'WINTER WEATHER' ~ 'BLIZZARD',
                EVTYPE == 'WINTER WEATHER/MIX' ~ 'BLIZZARD'
                
        ))

lookup$newEventType<-ifelse(is.na(lookup$newEventType),lookup$EVTYPE,lookup$newEventType)


#Merge new consolidated values into both dataframes
hhData<-merge(hhData,lookup, on='EVTYPE')
pdData<-merge(pdData,lookup, on='EVTYPE')






```


###Calculate human health impact likelihood, and average property damage per event type.

```{r calc human health likelihood}
#Summarize based on reduced event types and calculate likelihood based on number of effects per event
hhResultsData<-hhData %>% 
        group_by(newEventType) %>% 
        summarize(SumEvents = sum(NumberofEvents),
                  SumFatalies = sum(SumFatality),
                  SumInjuries = sum(SumInjury)
                  
                  ) %>% 
        mutate(
                InjuryLikelihood = round((SumInjuries/SumEvents),3),
               FatalityLikelihood = round((SumFatalies/SumEvents),3)
               )

pdResultsData<-pdData %>% 
        group_by(newEventType) %>% 
        summarize(SumEvents = sum(NumberofEvents),
                  SumPropertyDamages = sum(SumPropertyDamage)
                  
                  ) %>% 
        mutate(
                PropDamageAve = round((SumPropertyDamages/SumEvents),3)
                
        )
        



```


#Results

**Figure 1.** Summary of injury impacts by event type.
```{r figure 1}
gdata<-hhResultsData %>% select(newEventType, InjuryLikelihood) %>% arrange(InjuryLikelihood)

levs<-gdata$newEventType
gdata$newEventType<-factor(gdata$newEventType,levels=levs)


g<- ggplot(gdata,aes(newEventType,InjuryLikelihood))+
        geom_col(fill='orange')+
        coord_flip()+
        theme(axis.text.y = element_text(size=6))
g

```


**Figure 2.** Summary of fatality impacts by event type.

```{r figure 2}
gdata<-hhResultsData %>% select(newEventType, FatalityLikelihood) %>% arrange(FatalityLikelihood)

levs<-gdata$newEventType
gdata$newEventType<-factor(gdata$newEventType,levels=levs)


g<- ggplot(gdata,aes(newEventType,FatalityLikelihood))+
        geom_col(fill='wheat')+
        coord_flip()+
        theme(axis.text.y = element_text(size=6))
g
```




**Figure 3.** Summary of event type effect on property damage likelihood (dollars/event). (Top 50 event types shown out of 135)

```{r figure 3}
gdata<-pdResultsData %>% select(newEventType, PropDamageAve) %>% arrange(PropDamageAve)

levs<-gdata$newEventType
gdata$newEventType<-factor(gdata$newEventType,levels=levs)


g<- ggplot(gdata,aes(newEventType,PropDamageAve))+
        geom_col(fill='salmon')+
        coord_flip()+
        theme(axis.text.y = element_text(size=6))
g
```


















